Algorithms for Longest Common Extensions

### 2011-07-17 doc/report.tex
m

The Longest Common Extension problem can be used in many algorithms for solving other algorithmic problems, e.g.\ the Landau-Vishkin algorithm for approximate string searching~\cite{approx-search}. Optimal solutions exists for the problem, with constant query time and linear space. These teoretically optimal solutions are however not the best in practice, since they have large constant factors for both time and space usage. In average cases, a much simpler solution with worst case linear time, average case constant time, and no preprocessing has significantly better practical performance. This algorithm is ideal when only average case performance is relevant. In situations where we need both average case and worst case performance to be good, none of the existing solutions are ideal.

We have found two new algortihms. They both have better than linear worst case query times, and achieves significantly better average case practical query times compared to the theoretically best algorithms. The best of these two uses string fingerprinting to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for $k$ between one and $\log n$.

In \pref{sec:algorithms} we theorectically describe and analyze the LCE algorithms introduced here in \pref{sec:intro}, and in \pref{sec:results} we analyze results from practical C++ implementations of these algorithms.

@@

The algorithm stores LCE values for some of the $n^2$ possible query inputs on a given string. It uses \proc{DirectComp} to move towards one of these stored values, and stops when it reaches one. The \proc{DirectCompLookup} algorithm is very simple, but the tradoff between worst case query time and space is not very favorable, and preprocessing the data structure takes $O(n^2)$ time unless we use another LCE algorithm in the preprocessing.

@@

\paragraph{Tree notation.}
The ancestors of a node $u$ is $u$ itself as well as any node, which is a parent of an ancestor of $u$. The depth of a node $u$ is one greather than the depth of $u$'s parent, and the depth of the root is zero. A path in a tree begins at node $u$ and ends at node $v$, where $u$ is an ancestor of $v$.

@@

There are $n$ leaves in a suffix tree, and since each internal node has at least two children, the tree has less than $2n$ nodes. Each edge uses $O(1)$ space, as the label is a substring of $s$ and it can therefore be reprecented by two indexes of $s$. The suffix tree therefore takes $O(n)$ space.

@@

\paragraph{Predecessor.}
Given a set of $n$ objects, where the key of each object is a natural number between one and $U$, the predecessor of a natual number $x$ is an object with maximal key less than or equal to $x$. We can solve the static Predecessor problem in $O(\log\log U)$ time using $O(n)$ space~\cite{predecessor}.

\paragraph{Level Ancestor.}
The Level Ancestor at level $\ell$ of node $u$, $\LA(\ell,u)$, is the ancestor of $u$, whoes depth is $\ell$. We can preprocess a static tree in $O(n)$ time and space to answer Level Ancestor queries in $O(1)$ time~\cite{level-ancestor}.

@@

@@

For a string of length $n$ with aplhabet size $\sigma$, we can preprocess \proc{SuffixNca} in $O(\sortt)$ time by building the suffix tree, preprocessing it for NCA and building the array of label depths. The query time is $O(1)$, since looking up $L_i$ and $L_j$, doing an NCA query, and looking up the label depth in $D$ all takes constant time.

@@

Different algorithms exists for solving the RMQ problem, and we can solve it in optimal $O(1)$ time and $O(n)$ space. But like for the LCE problem, the asymptotically best RMQ solution may not be the best RMQ solution in practice. We examine the following RMQ solutions, where \RMQpq{p}{q} has $O(p)$ space and preprocessing time, and $O(q)$ query time:
\begin{samepage}
\begin{description}
\item[\RMQn] walks through the array directly in $O(n)$ time and does not use any preprocessing (except for transforming LCE into RMQ).
\item[\RMQq] preprocesses the array in $O(n)$ space in order to do queries in $O(1)$ time. This is a twolevel data structure as shown in \pref{fig:sa+lcp+min-twolevel}. The array is split into blocks of size $O(\log n)$. The top level is an array of length $O(n/\log n)$, where each element contains the minimum value one block. The top level uses a \RMQpq{n\log n}{1}-solution, and the lower level precomputes all answers for each block, which is done in $O(n)$ space, since there are only a limited number of different block types of size $O(\log n)$.
\item[\RMQlog] is the same as \RMQq\ except that the lower level is replaced with a direct seach through a $O(\log n)$ length block.
\end{description}
\end{samepage}

@@

The amount if I/O's used by \fprintk\ is $O(k\cdot n^{1/k})$. However if we structure our tables of fingerprints differently, we can improve ths number of I/O's to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model. Instead of storing the fingerprint of $s[i\twodots i+t_\ell]$ at $H_\ell[i]$, we can store it at $H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1]$. This will group all used fingerprints at level $\ell$ next to each other in memory, such that the amount of I/O's at each level is reduced from $O(n^{1/k})$ to $O(n^{1/k}/B)$.

@@

For larger values of $k$, we can attempt to optimize memory access accros different levels. When we move down through the levels for $k=\logceil$, we can continue in two directions after reading $H_\ell[i]$: $H_{\ell-1}[i]$ or $H_{\ell-1}[i+t_\ell]$. Neither of these positions are next to $H_\ell[i]$ in memory. We can optimize for the case where we reach $H_{\ell-1}[i]$ by flipping the direction in which we store the fingerprints, such that all fingerprints $H_\ell[i]$ for all $\ell$ and a specific $i$ are stored next to each other. This will not improve the asymptotic I/O bound, but could impove practical query times.

@@

We implemented different variants of \proc{DirectComp}, \proc{LcpRmq}, \proc{DirectCompLookup} and \fprintk\ in C++ and compared them to each other. We have not implemented \proc{SuffixNca}, since others have shown that \proc{LcpRmq} is better than \proc{SuffixNca} in practice~\cite{ilie-navarro-tinta}. The algorithms we compared are the following:
\begin{description}
\item[\proc{DirectComp}] is the simple \proc{DirectComp} algorithm with no preprocessing and worst case $O(n)$ query time.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac] is our new \fprintk\ algorithm using $k$ levels, where $k$ is $2$, $3$ and $\logceil$. The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater\ describe the exact size of fingerprinted substrings at each level.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater bc] is a variant of our new \fprintk\ algorithm, which starts at level $\ell=k-1$ instead of $\ell=0$, as discussed in \pref{sec:improved-avg}.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac\textless$u_0$, ..., $u_{k-2}$\textgreater] is a variant of our new \fprintk\ algorithm, which stays at level $\ell$ until $v\geq u_\ell$, as discussed in \pref{sec:improved-avg}, instead of doing one comparison at each level in step 1 of the query in \pref{sec:fingerprint-query}.
\item[\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{LcpRmq} algorithm using constant time RMQ.
\item[\proc{RMQ}\textless$p$, $q$\textgreater$_{\textrm{virtual}}$] is the \proc{LcpRmq} algorithm using different RMQ algorithms. They all contain an additional virtual method call to make implementation simpler, but their results cannot be directly compared to other algorithms, which do not have this extra virtual method call.
\item[\proc{DirectComp}\textless$c$\textgreater\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{DirectComp+LcpRmq} algorithm, where the cutoff from \proc{DirectComp} to \proc{LcpRmq} is after $c$ iterations.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac$_{\textrm{cache-horizontal-}\textit{op}}$] is a cache optimized variant of our \fprintk\ algorithm, where \textit{op} describes which of the shift or multiplication arithmetic operations it uses.
\end{description}

@@

We want to compare our new algortithms against the algortihm which is the best in practice of the asymptotically best $O(1)$ algorithms. The \proc{LcpRmq} algorithm is better than \proc{SuffixNca}~\cite{ilie-navarro-tinta}, so we only look at \proc{LcpRmq}. We have looked at a number of RMQ algorithms to find out which works best for the LCE problem. The results are shown in \pref{fig:plot-rmq}. \RMQq\ is the best in practice, and we therefore use that when we test against other LCE algorithms. This result is different from that of Ilie~et~al.~\cite{ilie-navarro-tinta}, where \RMQpq{n}{4\log n} is the best. This could be due to hardware or compiler differences.

@@

\pref{fig:plot-complookup} shows \proc{DirectCompLookup} with different time/space-tradeoffs together with \proc{LcpRmq}. In average case, \proc{DirectCompLookup} is slower than \proc{DirectComp}, most likely because it needs to do an extra ckeck to find out if it has reached a position for which a result is stored. For short strings, \proc{DirectLookup} is faster than \proc{DirectCompLookup}, since it does not need to check if the value it needs is stored. However, it quickly grows to be as slow as \proc{LcpRmq}, while \proc{DirectCompLookup} does not experience this slowdown. We explain this with that \proc{DirectLookup} reads from its $O(n^2)$ table of results in every query, whereas \proc{DirectCompLookup} only reads from the $O(n)$ string in average case, and thus has a significantly smaller working set.

@@

On worst case strings, the \fprintk\ algorithms are signicicantly better than \proc{DirectComp} and somewhat worse than \proc{LcpRmq}. Up until $n=30.000$ the three measured \fprintk\ algorithms have nearly the same query times. Of the \fprintk\ algorithms, the $k=2$ variant has a slight advantage for small strings of length less than around $2.000$. For longer strings the $k=3$ variant performs the best up to strings of length $250.000$, at which point the $k=\logceil$ variant becomes the best. This indicates that for shorter strings, using fewer levels is better, and when the input size increases, the \fprintk\ variants with better asymptotic query times have better worst case times in practice.

On strings with medium aveage LCE values, we see that our \fprintk\ algorithms are faster than both \proc{DirectComp} and \proc{LcpRmq}

We conclude that our new \fprintk\ algorithm acieves a tradeoff between worst case times and average case times, which is better than the existing best \proc{DirectComp} and \proc{LcpRmq} algorithms, yet it is not strictly better than the existing algorithms on all inputs. \fprintk\ is therefore a good choice in cases where both average case and worst case performance is important.

@@

In \pref{fig:plot-fingerprint-avglog} we see that on worst case strings, the query time of \fprint[\logceil] degrade a small amount when using the aveage case optimization. Both variations of \fprint[\logceil] show query times, whic are $O(\log n)$ as expected. On average case strings, the query time of \fprint[\logceil] improves significantly when using average case optimization, and both variants show their expected $O(1)$ and $O(\log n)$ average case query times.

@@

We conclude that using average case optimization yields a query time improvement on average case strings, which is significantly greater than the query time degration on worst case strings, and that \fprintk\ therefore should use the average case optimization.

@@

On average case strings the cache optimization does not change the query times, while on worst case strings and strings with medium size LCE values, cache optimization gives a noticable improvement for large inputs. The cache optimized \fprint[3] variant with shift operations shows an increase in query times for large inputs, which we cannot explain.

@@

Our new \fprintk\ algorithm is able to achieve a balance between worst case and average case query times. It has almost as good average case query times as \proc{DirectComp}, its worst case query times are significantly better than those of \proc{DirectComp}, and we have found cases between average and worst case where \fprintk\ is better than both \proc{DirectComp} and \proc{LcpRmq}. \fprintk\ gives a good time space tradeoff, and it uses less space than \proc{LcpRmq} when $k$ is small. The perfprmance of \fprintk\ can be tweaked on many parameters, and we have found that optimizing for average case queries and for memory caches improve practical query time performance.

@@

To test practical performance, we have implemented the algorithms in C++. The implementation is based on that of Ilie~et~al.~\cite{ilie-navarro-tinta}, which we have heavily modified. We have made it compile on a 32-bit Windows archictecture, and generalized it to allow more algorithms and tests to be implemented. \pref{tab:test-variables} summarizes the structure of the code and the parameters we can vary.

\begin{table}[tp]
\centering
\begin{tabular}{l l l}
Variable & Example & Component \\ \hline
alphabet size & $\sigma = 10$ & string \\
character pattern & uniformly at random & string \\ \hline
algorithm & \fprintk & algorithm \\
alg. parameter & $t = 2\sqrt n$ & algorithm \\ \hline
string length & exponentially increasing from $1$ to $100.000$ & test \\
query pattern & uniformly at random & test \\ \hline
\end{tabular}
\caption{List of performance veriables tested.}\label{tab:test-variables}
\end{table}

In our implementation, we have made the following assumptions about the LCE problem:
\begin{itemize}
\item A character in a string is always eight bits.
\item A string always ends with a special null character, which does not occur anywehere else in the string. This does not affect our results, since we do not measure practical preprocessing time. If an algorithm has faster query times when the string ends in a special character, it could just add that character to the end of the string in its preprocessing.
\item We make no assumptions on the order of $i$ and $j$. In some applications of the LCE problem, for example approximate string searching, we know that $i<j$, and some algorithms might be able to take advantage of this knowledge to get faster query times. We do not measure this case.
\item Wehenever we use random characters or query pairs, we use the \texttt{rand()} function from C. This gives repeatable pseudo-random numbers, which makes it easier for us to compare test runs.
\end{itemize}

@@

As an alternative way of testing our alorithms, we have looked at how long a query takes in relation to the LCE value it returns, since most of our algorithms have query times, which depend on the returned LCE value. \pref{fig:plot-by-value} shows query times for a fixed string of 1000 characters. We can see how the query times of \fprint[2] and \fprint[3] increases as the LCE value increases, and drops when the LCE value becomes long enough to make one more step at the higher level with fingerprints of longer substrings. We also see that the query time of \fprint[\log n] increases whenever the LCE value becomes long enough to reach a higher level of the data structure.

@@

We need to measure a sum of many queries to get an accurate timing, since each individual query is too fast to measure. While measureing a sequence of randomly generated queries is a very artificial test, which is unlikely to occur in practice, we think that it is even more artificial to measure a sequence of queries for which we already know that the LCE value is the same each time. On our worst case string $s=aaaaaaa...$, the queries we measure in each point in \pref{fig:plot-by-value} have the pattern $\max(i,j)=v$, which may give very different performance characteristics compared to random queries due to cache effects. We cannot meaningfully measure performance by LCE value on average case strings, as these strings do not contain large LCE values. The second plot in \pref{fig:plot-by-value} is on a string, which attempts to avoid a strong pattern like $\max(i,j)=v$ while it still has large LCE values.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{LZ78 definition\label{sec:lz-definition}}
A string $s$ of length $N$ has a corresponding \textit{compressed string} $z$ of compressed length $n$, where $z$ is a list of \textit{compression elements}, $z = z_1, z_2, ..., z_n$. Each compression element $z_x$ is a pair $(r_x, \alpha_x)$, where the reference, $\zreference_z(x)=r_x$, is an index of an earlier compression element in the compressed string or zero, i.e. $0 \leq \zreference_z(x) < x$, and the label, $\zlabel_z(x)=\alpha_z$, is a character from the original uncompressed string $s$. The \zphrase\ of a compression element is defined as:
\begin{align*}
\zphrase_z(x) =
\begin{cases}
    \zphrase_z(\zreference_z(x)) \cdot \zlabel_z(x) & \textrm{if}~ \zreference \neq 0\\
    \zlabel_z(x) & \textrm{otherwise}
\end{cases}
\end{align*}
where "$\cdot$" denotes string contatenation. From a compressed string $z = z_1,...,z_n$, we can get the original string $s=\zphrase_z(1)\cdot...\cdot\zphrase_z(n)$.

We can easily get the decompressed string $s$ from the compressed string $z$ using the definitions. To get the compressed string $z$ from the original string $s$ we can look at the compression elements as a tree, where the reference of each compression element corresponds to the parent pointer of each node in the tree. We consume $s$ from left to right while we build the tree:
\begin{enumerate}
\item Let the tree be single special root node $z_0$, let $i = 1$ and $c=z_0$.
\item If $c$ has a child $u$, where $\zlabel(u)=s[i]$, set $c=u$. Otherwise create the child node $u$, set $\zlabel(u)=s[i]$ and set $c=z_0$.
\item Increment $i$ by one and repeat from step 2 until the entire string is consumed when $i = n$.
\end{enumerate}
The zompressed string $z$ is now the tree except the special root node $z_0$.

@@

We can adapt \proc{DirectComp} to work on a compressed string. To work directly on a compressed string, we first need to have ramdom access to the characters of the original string.

@@

Since we have not been able to find improvements to the $O(N)$ worst case query time in $O(n)$ space using existing algorithms, we have looked into new algorithms. The compression elements of a compressed string $z$ forms a tree, and each phrase in $z$ is a path in the tree. The LCE value is the length of a substring occuring at two positions $i$ and $j$ in the input string. This substring spans a number of phrases, where each phrase is a path in the tree of compression elements. If we can make fast LCE queries on such a tree, we can make fast LCE queries on highly compressable strings.

@@

We could lay out the tree of compression elements in memory as a B-tree, where we store groups of adjacant nodes next to eachother in memory, to optimize the number of I/O's within each phrase, but each node may have more than $B$ children, and in this case it would not improve the worst case number of I/O's, as each node visited may be in a new group.

@@

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.7\textwidth,page=1]{tree-fingerprint.pdf}
    \end{center}
    \caption{\label{fig:tree-query}Finging the right fingerprint to look at for a \tfprintk\ query.}
\end{figure}

>>>>

The Longest Common Extension problem can be used in many algorithms for solving other algorithmic problems, e.g.\ the Landau-Vishkin algorithm for approximate string searching~\cite{approx-search}. Optimal solutions exists for the problem, with constant query time and linear space. These theoretically optimal solutions are however not the best in practice, since they have large constant factors for both time and space usage. In average cases, a much simpler solution with worst case linear time, average case constant time, and no preprocessing has significantly better practical performance. This algorithm is ideal when only average case performance is relevant. In situations where we need both average case and worst case performance to be good, none of the existing solutions are ideal.

We have found two new algorithms. They both have better than linear worst case query times, and achieves significantly better average case practical query times compared to the theoretically best algorithms. The best of these two uses string fingerprinting to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for $k$ between one and $\log n$.

In \pref{sec:algorithms} we theoretically describe and analyze the LCE algorithms introduced here in \pref{sec:intro}, and in \pref{sec:results} we analyze results from practical C++ implementations of these algorithms.

@@

The algorithm stores LCE values for some of the $n^2$ possible query inputs on a given string. It uses \proc{DirectComp} to move towards one of these stored values, and stops when it reaches one. The \proc{DirectCompLookup} algorithm is very simple, but the tradeoff between worst case query time and space is not very favorable, and preprocessing the data structure takes $O(n^2)$ time unless we use another LCE algorithm in the preprocessing.

@@

\paragraph{Tree notation.}
The ancestors of a node $u$ is $u$ itself as well as any node, which is a parent of an ancestor of $u$. The depth of a node $u$ is one greater than the depth of $u$'s parent, and the depth of the root is zero. A path in a tree begins at node $u$ and ends at node $v$, where $u$ is an ancestor of $v$.

@@

There are $n$ leaves in a suffix tree, and since each internal node has at least two children, the tree has less than $2n$ nodes. Each edge uses $O(1)$ space, as the label is a substring of $s$ and it can therefore be represented by two indexes of $s$. The suffix tree therefore takes $O(n)$ space.

@@

\paragraph{Predecessor.}
Given a set of $n$ objects, where the key of each object is a natural number between one and $U$, the predecessor of a natural number $x$ is an object with maximal key less than or equal to $x$. We can solve the static Predecessor problem in $O(\log\log U)$ time using $O(n)$ space~\cite{predecessor}.

\paragraph{Level Ancestor.}
The Level Ancestor at level $\ell$ of node $u$, $\LA(\ell,u)$, is the ancestor of $u$, whose depth is $\ell$. We can preprocess a static tree in $O(n)$ time and space to answer Level Ancestor queries in $O(1)$ time~\cite{level-ancestor}.

@@

\paragraph{Ladder Decomposition}
To make a ladder decomposition of a tree, we first create a ladder of nodes from the root of the tree to the leaf of greatest depth. We then remove the nodes in the ladder from the tree and repeat this operation recursively for all remaining subtrees, until all nodes are part of a ladder. Node $u$'s ladder is the ladder which contains $u$ at this point. Finally, we double the number of nodes in each ladder by extending it towards the root of the tree. If we reach the root before we have doubled the number of nodes, we stop at the root. Each node may thus be in multiple ladders.

For any path in a tree of $n$ nodes, we can find $O(\log n)$ ladders, which contain all the nodes in the path. If a node $u$ has height $h$, $u$'s ladder contains at least the last $h$ ancestors of $u$. The combined size of the ladders is at most $2n$.~\cite{level-ancestor}

@@

For a string of length $n$ with alphabet size $\sigma$, we can preprocess \proc{SuffixNca} in $O(\sortt)$ time by building the suffix tree, preprocessing it for NCA and building the array of label depths. The query time is $O(1)$, since looking up $L_i$ and $L_j$, doing an NCA query, and looking up the label depth in $D$ all takes constant time.

@@

Different algorithms exists for solving the RMQ problem, and we can solve it in optimal $O(1)$ time and $O(n)$ space. But like for the LCE problem, the asymptotically best RMQ solution may not be the best RMQ solution in practice. We examine the following RMQ solutions, where \RMQpq{p}{q} has $O(p)$ space and preprocessing time, and $O(q)$ query time:
\begin{samepage}
\begin{description}
\item[\RMQn] walks through the array directly in $O(n)$ time and does not use any preprocessing (except for transforming LCE into RMQ).
\item[\RMQq] preprocesses the array in $O(n)$ space in order to do queries in $O(1)$ time. This is a twolevel data structure as shown in \pref{fig:sa+lcp+min-twolevel}. The array is split into blocks of size $O(\log n)$. The top level is an array of length $O(n/\log n)$, where each element contains the minimum value one block. The top level uses a \RMQpq{n\log n}{1}-solution, and the lower level precomputes all answers for each block, which is done in $O(n)$ space, since there are only a limited number of different block types of size $O(\log n)$.
\item[\RMQlog] is the same as \RMQq\ except that the lower level is replaced with a direct search through a $O(\log n)$ length block.
\end{description}
\end{samepage}

@@

The amount if I/O's used by \fprintk\ is $O(k\cdot n^{1/k})$. However if we structure our tables of fingerprints differently, we can improve the number of I/O's to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model. Instead of storing the fingerprint of $s[i\twodots i+t_\ell]$ at $H_\ell[i]$, we can store it at $H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1]$. This will group all used fingerprints at level $\ell$ next to each other in memory, such that the amount of I/O's at each level is reduced from $O(n^{1/k})$ to $O(n^{1/k}/B)$.

@@

For larger values of $k$, we can attempt to optimize memory access accross different levels. When we move down through the levels for $k=\logceil$, we can continue in two directions after reading $H_\ell[i]$: $H_{\ell-1}[i]$ or $H_{\ell-1}[i+t_\ell]$. Neither of these positions are next to $H_\ell[i]$ in memory. We can optimize for the case where we reach $H_{\ell-1}[i]$ by flipping the direction in which we store the fingerprints, such that all fingerprints $H_\ell[i]$ for all $\ell$ and a specific $i$ are stored next to each other. This will not improve the asymptotic I/O bound, but could improve practical query times.

@@

We implemented different variants of \proc{DirectComp}, \proc{LcpRmq}, \proc{DirectCompLookup} and \fprintk\ in C++ and compared them to each other. We have not implemented \proc{SuffixNca}, since others have shown that \proc{LcpRmq} is better than \proc{SuffixNca} in practice~\cite{ilie-navarro-tinta}. The algorithms we compared are the following:
\begin{description}
\item[\proc{DirectComp}] is the simple \proc{DirectComp} algorithm with no preprocessing and worst case $O(n)$ query time.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac] is our new \fprintk\ algorithm using $k$ levels, where $k$ is $2$, $3$ and $\logceil$. The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater\ describe the exact size of fingerprinted substrings at each level.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater wc] is a variant of our new \fprintk\ algorithm, which starts at level $\ell=k-1$ instead of $\ell=0$, as discussed in \pref{sec:improved-avg}.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac\textless$u_0$, ..., $u_{k-2}$\textgreater] is a variant of our new \fprintk\ algorithm, which stays at level $\ell$ until $v\geq u_\ell$, as discussed in \pref{sec:improved-avg}, instead of doing one comparison at each level in step 1 of the query in \pref{sec:fingerprint-query}.
\item[\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{LcpRmq} algorithm using constant time RMQ.
\item[\proc{RMQ}\textless$p$, $q$\textgreater$_{\textrm{virtual}}$] is the \proc{LcpRmq} algorithm using different RMQ algorithms. They all contain an additional virtual method call to make implementation simpler, but their results cannot be directly compared to other algorithms, which do not have this extra virtual method call.
\item[\proc{DirectComp}\textless$c$\textgreater\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{DirectComp+LcpRmq} algorithm, where the cutoff from \proc{DirectComp} to \proc{LcpRmq} is after $c$ iterations.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac$_{\textrm{cache-horizontal-}\textit{op}}$] is a cache optimized variant of our \fprintk\ algorithm, where \textit{op} describes which of the shift or multiplication arithmetic operations it uses.
\end{description}

@@

We want to compare our new algorithms against the algorithm which is the best in practice of the asymptotically best $O(1)$ algorithms. The \proc{LcpRmq} algorithm is better than \proc{SuffixNca}~\cite{ilie-navarro-tinta}, so we only look at \proc{LcpRmq}. We have looked at a number of RMQ algorithms to find out which works best for the LCE problem. The results are shown in \pref{fig:plot-rmq}. \RMQq\ is the best in practice, and we therefore use that when we test against other LCE algorithms. This result is different from that of Ilie~et~al.~\cite{ilie-navarro-tinta}, where \RMQpq{n}{4\log n} is the best. This could be due to hardware or compiler differences.

@@

\pref{fig:plot-complookup} shows \proc{DirectCompLookup} with different time/space-tradeoffs together with \proc{LcpRmq}. In average case, \proc{DirectCompLookup} is slower than \proc{DirectComp}, most likely because it needs to do an extra check to find out if it has reached a position for which a result is stored. For short strings, \proc{DirectLookup} is faster than \proc{DirectCompLookup}, since it does not need to check if the value it needs is stored. However, it quickly grows to be as slow as \proc{LcpRmq}, while \proc{DirectCompLookup} does not experience this slowdown. We explain this with that \proc{DirectLookup} reads from its $O(n^2)$ table of results in every query, whereas \proc{DirectCompLookup} only reads from the $O(n)$ string in average case, and thus has a significantly smaller working set.

@@

On worst case strings, the \fprintk\ algorithms are significantly better than \proc{DirectComp} and somewhat worse than \proc{LcpRmq}. Up until $n=30.000$ the three measured \fprintk\ algorithms have nearly the same query times. Of the \fprintk\ algorithms, the $k=2$ variant has a slight advantage for small strings of length less than around $2.000$. For longer strings the $k=3$ variant performs the best up to strings of length $250.000$, at which point the $k=\logceil$ variant becomes the best. This indicates that for shorter strings, using fewer levels is better, and when the input size increases, the \fprintk\ variants with better asymptotic query times have better worst case times in practice.

On strings with medium average LCE values, we see that our \fprintk\ algorithms are faster than both \proc{DirectComp} and \proc{LcpRmq}

We conclude that our new \fprintk\ algorithm achieves a tradeoff between worst case times and average case times, which is better than the existing best \proc{DirectComp} and \proc{LcpRmq} algorithms, yet it is not strictly better than the existing algorithms on all inputs. \fprintk\ is therefore a good choice in cases where both average case and worst case performance is important.

@@

In \pref{fig:plot-fingerprint-avglog} we see that on worst case strings, the query time of \fprint[\logceil] degrade a small amount when using the average case optimization. Both variations of \fprint[\logceil] show query times, which are $O(\log n)$ as expected. On average case strings, the query time of \fprint[\logceil] improves significantly when using average case optimization, and both variants show their expected $O(1)$ and $O(\log n)$ average case query times.

@@

We conclude that using average case optimization yields a query time improvement on average case strings, which is significantly greater than the query time increase on worst case strings, and that \fprintk\ therefore should use the average case optimization.

@@

On average case strings the cache optimization does not change the query times, while on worst case strings and strings with medium size LCE values, cache optimization gives a noticeable improvement for large inputs. The cache optimized \fprint[3] variant with shift operations shows an increase in query times for large inputs, which we cannot explain.

@@

Our new \fprintk\ algorithm is able to achieve a balance between worst case and average case query times. It has almost as good average case query times as \proc{DirectComp}, its worst case query times are significantly better than those of \proc{DirectComp}, and we have found cases between average and worst case where \fprintk\ is better than both \proc{DirectComp} and \proc{LcpRmq}. \fprintk\ gives a good time space tradeoff, and it uses less space than \proc{LcpRmq} when $k$ is small. The performance of \fprintk\ can be tweaked on many parameters, and we have found that optimizing for average case queries and for memory caches improve practical query time performance.

@@

To test practical performance, we have implemented the algorithms in C++. The implementation is based on that of Ilie~et~al.~\cite{ilie-navarro-tinta}, which we have heavily modified. We have made it compile on a 32-bit Windows architecture, and generalized it to allow more algorithms and tests to be implemented. \pref{tab:test-variables} summarizes the structure of the code and the parameters we can vary.

\begin{table}[tp]
\centering
\begin{tabular}{l l l}
Variable & Example & Component \\ \hline
alphabet size & $\sigma = 10$ & string \\
character pattern & uniformly at random & string \\ \hline
algorithm & \fprintk & algorithm \\
alg. parameter & $t = 2\sqrt n$ & algorithm \\ \hline
string length & exponentially increasing from $1$ to $100.000$ & test \\
query pattern & uniformly at random & test \\ \hline
\end{tabular}
\caption{List of performance variables tested.}\label{tab:test-variables}
\end{table}

In our implementation, we have made the following assumptions about the LCE problem:
\begin{itemize}
\item A character in a string is always eight bits.
\item A string always ends with a special null character, which does not occur anywhere else in the string. This does not affect our results, since we do not measure practical preprocessing time. If an algorithm has faster query times when the string ends in a special character, it could just add that character to the end of the string in its preprocessing.
\item We make no assumptions on the order of $i$ and $j$. In some applications of the LCE problem, for example approximate string searching, we know that $i<j$, and some algorithms might be able to take advantage of this knowledge to get faster query times. We do not measure this case.
\item Whenever we use random characters or query pairs, we use the \texttt{rand()} function from C. This gives repeatable pseudo-random numbers, which makes it easier for us to compare test runs.
\end{itemize}

@@

As an alternative way of testing our algorithms, we have looked at how long a query takes in relation to the LCE value it returns, since most of our algorithms have query times, which depend on the returned LCE value. \pref{fig:plot-by-value} shows query times for a fixed string of 1000 characters. We can see how the query times of \fprint[2] and \fprint[3] increases as the LCE value increases, and drops when the LCE value becomes long enough to make one more step at the higher level with fingerprints of longer substrings. We also see that the query time of \fprint[\log n] increases whenever the LCE value becomes long enough to reach a higher level of the data structure.

@@

We need to measure a sum of many queries to get an accurate timing, since each individual query is too fast to measure. While measuring a sequence of randomly generated queries is a very artificial test, which is unlikely to occur in practice, we think that it is even more artificial to measure a sequence of queries for which we already know that the LCE value is the same each time. On our worst case string $s=aaaaaaa...$, the queries we measure in each point in \pref{fig:plot-by-value} have the pattern $\max(i,j)=v$, which may give very different performance characteristics compared to random queries due to cache effects. We cannot meaningfully measure performance by LCE value on average case strings, as these strings do not contain large LCE values. The second plot in \pref{fig:plot-by-value} is on a string, which attempts to avoid a strong pattern like $\max(i,j)=v$ while it still has large LCE values.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{LZ78 definition\label{sec:lz-definition}}
A string $s$ of length $N$ has a corresponding \textit{compressed string} $z$ of compressed length $n$, where $z$ is a list of \textit{compression elements}, $z = z_1, z_2, ..., z_n$. Each compression element $z_x$ is a pair $(r_x, \alpha_x)$, where the reference, $\zreference_z(x)=r_x$, is an index of an earlier compression element in the compressed string or zero, i.e. $0 \leq \zreference_z(x) < x$, and the label, $\zlabel_z(x)=\alpha_z$, is a character from the original uncompressed string $s$. The \zphrase\ of a compression element is defined as:
\begin{align*}
\zphrase_z(x) =
\begin{cases}
    \zphrase_z(\zreference_z(x)) \cdot \zlabel_z(x) & \textrm{if}~ \zreference \neq 0\\
    \zlabel_z(x) & \textrm{otherwise}
\end{cases}
\end{align*}
where "$\cdot$" denotes string concatenation. From a compressed string $z = z_1,...,z_n$, we can get the original string $s=\zphrase_z(1)\cdot...\cdot\zphrase_z(n)$.

We can easily get the decompressed string $s$ from the compressed string $z$ using the definitions. To get the compressed string $z$ from the original string $s$ we can look at the compression elements as a tree, where the reference of each compression element corresponds to the parent pointer of each node in the tree. We consume $s$ from left to right while we build the tree:
\begin{enumerate}
\item Let the tree be single special root node $z_0$, let $i = 1$ and $c=z_0$.
\item If $c$ has a child $u$, where $\zlabel(u)=s[i]$, set $c=u$. Otherwise create the child node $u$, set $\zlabel(u)=s[i]$ and set $c=z_0$.
\item Increment $i$ by one and repeat from step 2 until the entire string is consumed when $i = n$.
\end{enumerate}
The compressed string $z$ is now the tree except the special root node $z_0$.

@@

We can adapt \proc{DirectComp} to work on a compressed string. To work directly on a compressed string, we first need to have random access to the characters of the original string.

@@

Since we have not been able to find improvements to the $O(N)$ worst case query time in $O(n)$ space using existing algorithms, we have looked into new algorithms. The compression elements of a compressed string $z$ forms a tree, and each phrase in $z$ is a path in the tree. The LCE value is the length of a substring, which occurs at two positions $i$ and $j$ in the input string. This substring spans a number of phrases, where each phrase is a path in the tree of compression elements. If we can make fast LCE queries on such a tree, we can make fast LCE queries on highly compressable strings.

@@

We could lay out the tree of compression elements in memory as a B-tree, where we store groups of adjacent nodes next to each other in memory, to optimize the number of I/O's within each phrase, but each node may have more than $B$ children, and in this case it would not improve the worst case number of I/O's, as each node visited may be in a new group.

@@

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.7\textwidth,page=1]{tree-fingerprint.pdf}
    \end{center}
    \caption{\label{fig:tree-query}Finding the right fingerprint to look at for a \tfprintk\ query.}
\end{figure}
